### 2015-06-15 chapters/ipa_meets_unicode.tex
minor correction
updated diacritic ordering section and committing initial IPA profile withnames.tsv

Furthermore, even linguists are unlikely to distinguish between
the <ə> \textsc{latin small letter schwa} at code point \uni{0259} and <ǝ>
\textsc{latin small letter turned e} at \uni{01DD}.

@@

\begin{itemize}
	\item The uses of the apostrophe has led to long discussions on the Unicode
       Standard email list. An English keyboard inputs <{\fontspec{Monaco}'}> \textsc{apostrophe}
       at \uni{0027}, although the preferred Unicode apostrophe is the <\ ' >
       \textsc{right single quotation mark} at \uni{2019}.
     \item The glottal stop/glottalization/ejective marker is another completely
        different character <{\large ʼ}>, the \textsc{modifier letter apostrophe} at
        \uni{02BC}, but unfortunately looks mostly highly similar to
        \uni{2019}. 
	\item Another problem is the <ˁ> \textsc{modifier letter reversed
       glottal stop} at \uni{02C1} vs.\@ the <ˤ> \textsc{modifier
       letter small reversed glottal stop} at \uni{02E4}. Both 
       appear in various resources representing phonetic data online. This is
       thus a clear example for which the Unicode Standard does not solve the
       linguistic standardization problem. 
	\item There is at least one case in which the character name assigned by the
       Unicode Consortium does not match the IPA's description. In the Unicode
       Standard the <ǃ> at \uni{01C3} is labeled \textsc{latin letter retroflex
       click}, but in IPA that glyph is used for an alveolar or postalveolar
       click (not retroflex). This naming is probably best seen as a simple
       error in the Unicode Standard. This character is of course often simply
       typed as <!> \textsc{exclamation mark} at \uni{0021}. \end{itemize}

@@

Also related to diacritics is the question of ordering. To our knowledge, the
International Phonetic Association does not specify a specific ordering for
diacritics that combine with phonetic base symbols; this exercise is left to the
reasoning of the transcriber. However, such marks have to be explicitly ordered
if sequences of them are to be interoperable and compatible. An example is a
labialized aspirated alveolar plosive: <tʷʰ>. There is nothing holding linguists
back from using <tʰʷ> instead (with exactly the same intended meaning). However,
from a technical standpoint, these two sequences are different, e.g.~if both
sequences are used in a document, searching for <tʷʰ> will not find any
instances of <tʰʷ>, and vice versa. Likewise, a creaky voiced syllabic dental
nasal can be encoded in various orders, e.g. <n̪̰̩>, <n̩̰̪> or <n̩̪̰>.

@@

In accordance with the absence of any specification of ordering in the IPA, the
Unicode Standard likewise does not propose any standard orderings. Both leave it
to the user to be consistent. However, there is one aspect of ordering for which
the Unicode Standard does present a canonical solution, which is uncontroversial
from a linguistic perspective. Diacritics in the Unicode Standard
(i.e.~Combining Diacritical Marks, see above) are classified in Canonical
Combining Classes. In practice, the diacritics are distinguished by their
position relative to the base character.\footnote{See
\url{http://unicode.org/reports/tr44/\#Canonical\_Combining\_Class\_Values} for a
detailed description.} When applying a Unicode normalization (NFC or NFD, see
Section~\ref{pitfall-canonical-equivalence}), the diacritics in different
positions are put in a specified order. This process therefore harmonizes the
difference between different encodings, for example, of a midtone creaky voice
vowel <ḛ̄>. This grapheme cluster can be encoded either as <e> + <\ \ ̄> + <\ \
̰> or as <e> + <\ \ ̰> + <\ \ ̄>. To prevent this twofold encoding, the Unicode
Standard specifies the second ordering as canonical (in this case, diacritics
below are put before diacritics above).

@@

Facing the problem of specifying a consistent ordering of diacritics while
developing a large database of phonological inventories from the world's
languages, \citet[540]{Moran2012} defines a set of diacritic ordering
conventions.\footnote{The most recent version of these conventions is online at
\url{http://phoible.github.io/conventions/}} The conventions are influenced by
the linguistic literature, though some ad-hoc decisions had to be taken. The
goal was to explicitly define all character sequences so that the vast variety
of phonemes found in descriptions of the world's language were normalized into
consistent character sequences, e.g.~if one language description uses and
another, when both are intended to be phonetically equivalent, then a decision
to normalize to one form was taken. For example, when a character sequence
contains more than one character in Spacing Modifier Letters, the order that is
proposed is the following (where <\textbar{}> indicates \textit{or} and <→>
indicates \textit{precedes}):

\begin{itemize}
	\item[] \textsc{Spacing Modifier Letters ordering:}
	\begin{itemize}
	  \item[→] (unreleased <\ \ ̚ > \textbar{} lateral release <ˡ> \textbar{} nasal release <ⁿ>)
	  \item[→] (palatalized <ʲ>)
	  \item[→] (labialized <ʷ>)
	  \item[→] (velarized <ˠ>)
	  \item[→] (pharyngealized <ˤ>)
	  \item[→] (aspirated <ʰ> \textbar{} ejective <ʼ>)
	  \item[→] (long <ː> \textbar{} half-long <ˑ>)
	\end{itemize}
\end{itemize}

If a character sequence contains more than one diacritic below the base
character, then the place feature is applied first (dental, laminal, apical,
fronted, backed, lowered, raised), then the laryngeal setting (voiced,
voiceless, creaky voice, breathy voice) and finally the syllabic or non-syllabic
marker (for vowels, ATR gets put on between the place and laryngeal setting).
So:

\begin{itemize}
	\item[] \textsc{Combining Diacritical Marks (below) ordering:}
	\begin{itemize}	
	  \item[→] (dental <t̪> \textbar{} laminal <t̻> \textbar{} apical <t̺>) 
	  \item[→] (fronted <u̟> \textbar{} backed <e̠>)
	  \item[→] (lowered <e̞> \textbar{} raised <e̝>) 
	  \item[→] (ATR <e̘ e̙>)
	  \item[→] (voiced <s̬> \textbar{} voiceless <n̥> \textbar{} creaky voice <b̰> \textbar{} breathy voice <b̤>)
	  \item[→] (syllabic <n̩> \textbar{} non-syllabic <e̯>)
	\end{itemize}
 \end{itemize}

Character sequences with diacritics above the base character were not
problematic in \citet{Moran2012} because they include only the centralized,
mid-centralized and nasalized combining characters. \citet{Moran2012} marks
tones as singletons with Space Modifier Letters, e.g. \textless{}˦\textgreater{}
for a phonemic high tone, instead of accent diacritics, alleviating potential
conflicts. Building on the work of \citet{Moran2012}, if a character sequence
contains more than one diacritic above the base character, we propose:

>>>>

  \item The uses of the apostrophe has led to long discussions on the Unicode
     Standard email list. An English keyboard inputs <{\fontspec{Monaco}'}> \textsc{apostrophe}
     at \uni{0027}, although the preferred Unicode apostrophe is the <\ ' >
     \textsc{right single quotation mark} at \uni{2019}.
  \item The glottal stop/glottalization/ejective marker is another completely
      different character <{\large ʼ}>, the \textsc{modifier letter apostrophe} at
      \uni{02BC}, but unfortunately looks mostly highly similar to
      \uni{2019}. 
  \item Another problem is the <ˁ> \textsc{modifier letter reversed
     glottal stop} at \uni{02C1} vs.\@ the <ˤ> \textsc{modifier
     letter small reversed glottal stop} at \uni{02E4}. Both 
     appear in various resources representing phonetic data online. This is
     thus a clear example for which the Unicode Standard does not solve the
     linguistic standardization problem.
  \item even linguists are unlikely to distinguish between the <ə>
     \textsc{latin small letter schwa} at code point \uni{0259} and <ǝ>
     \textsc{latin small letter turned e} at \uni{01DD}.
  \item There is at least one case in which the character name assigned by the
     Unicode Consortium does not match the IPA's description. In the Unicode
     Standard the <ǃ> at \uni{01C3} is labeled \textsc{latin letter retroflex
     click}, but in IPA that glyph is used for an alveolar or postalveolar
     click (not retroflex). This naming is probably best seen as a simple
     error in the Unicode Standard. This character is of course often simply
     typed as <!> \textsc{exclamation mark} at \uni{0021}.

@@

@@

Also related to diacritics is the question of ordering. To our knowledge, the
International Phonetic Association does not specify a specific ordering for
diacritics that combine with phonetic base symbols; this exercise is left to the
reasoning of the transcriber. However, such marks have to be explicitly ordered
if sequences of them are to be interoperable and compatible computationally. An example is a
labialized aspirated alveolar plosive: <tʷʰ>. There is nothing holding linguists
back from using <tʰʷ> instead (with exactly the same intended meaning). However,
from a technical standpoint, these two sequences are different, e.g.~if both
sequences are used in a document, searching for <tʷʰ> will not find any
instances of <tʰʷ>, and vice versa. Likewise, a creaky voiced syllabic dental
nasal can be encoded in various orders, e.g. <n̪̰̩>, <n̩̰̪> or <n̩̪̰>.

@@

In accordance with the absence of any specification of ordering in the IPA, the
Unicode Standard likewise does not propose any standardized orders. Both leave it
to the user to be consistent; this approach naturally invites inconsistency across 
different authored resources.

There is one aspect of ordering for which
the Unicode Standard does present a canonical solution. However, it is uncontroversial
from a linguistic perspective. Diacritics in the Unicode Standard
(i.e.~Combining Diacritical Marks, see above) are classified in Canonical
Combining Classes. In practice, the diacritics are distinguished by their
position relative to the base character.\footnote{See
\url{http://unicode.org/reports/tr44/\#Canonical\_Combining\_Class\_Values} for a
detailed description.} When applying a Unicode normalization (NFC or NFD, see
Section~\ref{pitfall-canonical-equivalence}), the diacritics in different
positions are put in a specified order. This process therefore harmonizes the
difference between different encodings, e.g.\ of a midtone creaky voice
vowel <ḛ̄>. This grapheme cluster can be encoded either as <e> + <\ \ ̄> + <\ \
̰> or as <e> + <\ \ ̰> + <\ \ ̄>. To prevent this twofold encoding, the Unicode
Standard specifies the second ordering as canonical (in this case, diacritics
below are put before diacritics above).

@@

Facing the problem of specifying a consistent ordering of diacritics while
developing a large database of phonological inventories from the world's
languages, \citet[540]{Moran2012} defines a set of diacritic ordering
conventions. As new languages' phonological inventories are added to the 
database,\footnote{\url{http://phoible.org/}} updates and additions to the 
conventions are published online by \cite{MoranMcCloy2014}.\footnote{
\url{http://phoible.github.io/conventions/}} The conventions are influenced by
the linguistic literature, though some ad-hoc decisions had to be taken, like 
introducing several symbols for widely attested sounds that lack representation 
in the current version of the IPA. Whenever possible, additions were drawn from 
the extIPA symbols for disordered speech.\footnote{\url{https://www.internationalphoneticassociation.org/sites/default/files/extIPAChart2008.pdf}}

\cite{MoranMcCloy2014} explicitly define all character sequences so that the vast variety
of phonemes found in a large sample of descriptions of the world's language were normalized into
consistent character sequences, e.g.~if one language description uses <kʲʼ> and
another <kʼʲ>, when both are intended to be phonetically equivalent, then a decision
to normalize to one form was taken. For example, when a character sequence
contains more than one character in Spacing Modifier Letters, the order that is
proposed is the following (where <\textbar{}> indicates \textit{or} and <→>
indicates \textit{precedes}):

\begin{itemize}
	\item[] \textsc{Spacing Modifier Letters ordering:}
	\begin{itemize}
	  \item[→] (unreleased <\ \ ̚ >)
	  \item[→] (rhotic hook <\ \ ˞ >)
	  \item[→] (lateral release <ˡ> \textbar{} nasal release <ⁿ>)
	  \item[→] (labialized <ʷ>)
	  \item[→] (palatalized <ʲ>)
	  \item[→] (velarized <ˠ>)
	  \item[→] (pharyngealized <ˤ>)
	  \item[→] (glottalized <ˀ>)
	  \item[→] (aspirated <ʰ> \textbar{} ejective <ʼ>)
	  \item[→] (long <ː> \textbar{} half-long <ˑ>)
	\end{itemize}
\end{itemize}

If a character sequence contains more than one diacritic below the base
character, then the place features are applied first (linguolabial, dental, apical, laminal, advanced, 
retracted), followed by the manner features (raised, lowered, advanced tongue root, 
retracted tongue root, frictionalized), then secondary articulations (more round, less round, 
derhoticized), laryngeal settings (creaky, breathy, voiced, devoiced (below), devoiced (above)), and finally the syllabic or non-syllabic
marker (for vowels, advanced and retracted tongue root markers gets applied between the place and laryngeal setting).
So:

\begin{itemize}
	\item[] \textsc{Combining Diacritical Marks (below) ordering:}
	\begin{itemize}	
	  \item[→] (linguolabial <t̼> \textbar{} dental <t̪> \textbar{} apical <t̺> \textbar{} laminal <t̻>)
	  \item[→] (advanced <u̟> \textbar{} retracted <e̠>)	 
	  \item[→] (raised <e̝> \textbar{} lowered <e̞>) 
	  \item[→] (advanced tongue root <e̘> \textbar{} retracted tongue root <e̙>)
	  \item[→] (frictionalized <ɾ͓>)
	  \item[→] (more round <ɔ̹> \textbar{} less round <ɔ̜>)
	  \item[→] (creaky voice <b̰> \textbar{} breathy voice <b̤> \textbar{} voiced <s̬> \textbar{} devoiced <n̥>)
	  \item[→] (syllabic <n̩> \textbar{} non-syllabic <e̯>)
	\end{itemize}
 \end{itemize}

Character sequences with diacritics above the base character were not
problematic in \citet{Moran2012} because they include only the centralized,
mid-centralized and nasalized combining characters.

\citet{Moran2012,MoranMcCloy2014} mark
tones as singletons with Space Modifier Letters, e.g. \textless{}˦\textgreater{}
for a phonemic high tone, instead of accent diacritics, alleviating potential
conflicts. Building on these works, if a character sequence
contains more than one diacritic above the base character, we propose:
